---
title: "Course 3 - Getting and Cleaning Data - Week 3 - Notes"
author: "Greg Foletta"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir = '~/Documents/Study/Coursera_Data_Science_JHU/C3_Getting_and_Cleaning_Data/')
library(tidyverse)
```

# Summarising Data

```{r restaurants}
bmore_rstrnts <- read_csv('https://data.baltimorecity.gov/api/views/k5ry-ef3g/rows.csv?accessType=DOWNLOAD')
```
Take a quick glance at the data.
```{r looking}
# Look at the first few entries
head(bmore_rstrnts)
# Look at the last few entries.
tail(bmore_rstrnts)
# A summary of the data frame
summary(bmore_rstrnts)
```

We can look at the quantiles of quantitative variables.

```{r quantiles}
quantile(bmore_rstrnts$councilDistrict)
```
Can create a table. The `useNA` argument controls whether the table includes NA values. You can use:
- 'no' - don't include.
- 'ifany' - only of the count is positive.
- 'always' - even if there are zero counts.

The default is **no**.

```{r table}
table(bmore_rstrnts$zipCode, useNA = 'ifany')
```

Can use two variables:

```{r table_two_vars}
table(bmore_rstrnts$councilDistrict, bmore_rstrnts$zipCode)
```

## Check For Missing Values

```{r check_missing}
# How many missing values
sum(is.na(bmore_rstrnts$councilDistrict))

# Could also see if there are any missing values
any(is.na(bmore_rstrnts$councilDistrict))

# Are all the zipcodes greater than 0
all(bmore_rstrnts$zipCode > 0)
```

## Row and Column Sums

Check which columns have NA values

```{r column_nas}
colSums(is.na(bmore_rstrnts))
```

## Values With Specific Characteristics

```{r values}
table(bmore_rstrnts$zipCode %in% c('21212'))

head(bmore_rstrnts[bmore_rstrnts$zipCode %in% c('21212'), ])
```

## Cross Tabs

```{r cross_tabs}
UCBAdmissions %>% as.data.frame() %>% summary()

xtabs(Freq ~ Gender + Admit, data = UCBAdmissions)
```

## Flat Tables

With lots of variables the cross-tabulation becomes very hard to read. Can use `ftable()` instead.
```{r flat_tables}
warpbreaks$replicate <- rep(1:9, len = 54)
xtabs(breaks ~ ., data = warpbreaks)

head(ftable(warpbreaks))
```

## Size in Memory

Use `object.size()`.

```{r object_size}
big_vec <- 1:1000000000
object.size(big_vec)
print(object.size(big_vec), units = 'Mb')
```

# Creating New Variables

Often the raw data doesn't have a value you're looking for. You need to transform the data to get the values you would like.

Common variables:
- 'Missingness' indicators.
- 'Cutting Up' quantitative variables.
- Applying transforms.

## Creating Categorical Variables

Use `cut()` to cut up quantitative variables.

```{r cutting}
bmore_rstrnts$zipGroups <- cut(bmore_rstrnts$zipCode, breaks = quantile(bmore_rstrnts$zipCode))

table(bmore_rstrnts$zipGroups)
```

Can use `cut2()` from the `Hmisc` library to make it easier:

```{r cut2, message = F}
library(Hmisc)
bmore_rstrnts$zipGroups <- cut2(bmore_rstrnts$zipCode, g = 4)

table(bmore_rstrnts$zipGroups)
```

# Reshaping Data

You can melt the data frame, pulling the measurement variables into a single column:

```{r melt}
library(reshape2)
car_data <- mtcars %>%
    as_tibble() %>%
    rownames_to_column(var = 'carname')

car_melt <- car_data %>% melt(id = c('carname', 'gear', 'cyl'), measure.vars = c('mpg', 'hp'))
```

## Casting

The `dcast()` function allows you to recast the frame into a different shape. We see cylinders in terms of `mpg` and `hp`.

```{r casting}
# Standard
car_melt %>% dcast(cyl ~ variable)

# Can also pass a different function to summarise with
car_melt %>% dcast(cyl ~ variable,mean)
```

## Averaging Variables

Use `tapply()`:

```{r tapply_average}
tapply(InsectSprays$count, InsectSprays$spray, sum)
```

Can also use `split()`:

```{r split_average}
InsectSplits <- split(InsectSprays$count, InsectSprays$spray)
InsectSplits

InsectSplits <- lapply(InsectSplits, sum)
unlist(InsectSplits)
```

# Merging Data

The `merge()` function can merge data frames together. By default it merges by all of the columns that have a common name.


```{r merge}
a <- tibble(
    aa = 1:5,
    numbera_a = 6:10,
    normal_a = rnorm(5)
)

b <- tibble(
    aa = 1:5,
    number_b = 6:10,
    normal_b = rnorm(5)
)

merge(a, b)
merge(a, b, by = 'aa')
```

An alternative is to use `*_join()` functons from `dplyr`.

```{r dplyr_join}
full_join(a,b)
```

