---
title: "Course 3 - Getting and Cleaning Data - Week 4 - Notes"
author: "Greg Foletta"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir = '~/Documents/Study/Coursera_Data_Science_JHU/C3_Getting_and_Cleaning_Data/')
library(tidyverse)
```

# Fixing Character Vectors

Fix the case of character vectors using `tolower()` and `toupper()`.

Split up a word using `strsplit()`

```{r strsplit}
string_to_split <- c('this.1', 'something.2', 'another')
strsplit(string_to_split, '\\.')
```

Substitute out characters using the `sub()` function.

```{r sub}
strings_to_sub <- c('this-is-going-to', 'replace-some', 'hyphens')

# sub replaces only the first instance
sub('-', '_', strings_to_sub)

# gsub replaces all instances
gsub('-', '_', strings_to_sub)
```

Use `grep()` and `grepl()` to find values. `grep()` returns a vector of the matching indicies, and `grepl()` returns a logical vector of matches and not matches.

```{r grep}
greppable <- c('these', 'are', 'some', 'words', 'to', 'grep')

# Return indicies
grep('to', greppable)
# Return a logical vector
grepl('o', greppable)
#Return the values
grep('[swt]o', greppable, value = T)
```

# Working with Dates

## Dates

Date function:

```{r date}
d1 <- date()
d1
class(d1)

d2 <- Sys.Date()
d2
class(d2)
```

Use the `format()` function to alter the format of the date.

```{r format}
format(d2, "%A %d %B %Y")
```

To create a date, use the `as.Date()` function.

```{r asdate}
some_dates <- c('01oct1990', '02sep1991')
d3 <- as.Date(some_dates, '%d%b%Y')
weekdays(d3)
```

The `Lubridate` package is very good when working with dates / times.

```{r lubridate}
library(lubridate)

ymd('19200202')

dmy('01-01-2001')
```

## Times

Lubridate also has functions for working with times.

```{r lubridate_t}
ymd_hms('2002-01-01 14:01:01')
ymd_hms('2002-01-01 14:01:01', tz = 'Australia/Melbourne')
```
