---
title: "Course 9 - Developing Data Products- Week 1 - Notes"
author: "Greg Foletta"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
    toc: true
---


```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
```

# Shiny

Shiny is a web development framework in R. It allows you to create a graphical interface so that users can interact with visualisations, models and algorithms without needing to know R themselves.

# An Application

When you create an app, there's a ui.R and a server.R.

Within the UI, you start off with a `shinyUI()` call which one arugment:

- A `fluidPage()` call to set up the page. This function has two arguments:
    - A `titlePanel()` to set the title.
    - A `sidebarLayout()` to set the sidebar, and
    - A `mainPanel()` to set the main content.
    
```r
library(shiny)

# Define UI for application that draws a histogram
shinyUI(fluidPage(
    # Application title
    titlePanel("Old Faithful Geyser Data"),

    # Sidebar with a slider input for number of bins
    sidebarLayout(
        sidebarPanel( h3('Sidebar Text') ),
        mainPanel( h3("Main Panel Text') )
    )
))
```


# Wrapper Tags

Shiny provides several wrapper functions for suing standard HTML tags:

- `h1()` - `h6()`
- `p()`
- `a()`
- `div()`
- `span()`
- `code()`

Can look at `?builder` for more information.

# Inputs and Outputs

## Slider

A slider can be created as such on the UI side:

```r
sliderInpput('slider_id' 'Text', min, max, starting_value)
```

Then on the server side:

```r
shinyServer(function(input, output) {
    output$slider_text= renderText(input$slider_id)
})
```

Then back on the UI we can use

```r
textOutput('slider_text')
```

to output the value.

# Plots

Shiny provides the `plotOutput()` function and the `renderPlot()` for taking user input and creating plots

On the user side we use `plotOutput('plot_name')`. On the server side we use the render plot function. Here's an example:

```r
output$plot_name <- render_plot({
    set.seed(1)
    data_x <- rnorm(100, 20)
    data_y <- rnorm(100, 20)
    plot(data_x, data_y)
})
```

# Reactivity

A reactive expression manipulates inputs from Shiny and returns a value. It provides a way for an app to respond since inputs will change depending on how users interact with the user interface.

Expressions wrapped by `reactive()` are subject to change.

```r
calc_sum <- reactive({
    input$box1 + input$box2
})

calc_sum()
```

You may not want the app to immediately react to changes in user input, as the calculation may be long running. This is where delayed reactivity comes in.

You add a button in the UI and this delays the reactivity:

```r
submitButton('submit')
```


# Tabs

There are several other kinds of UI components including tabs, navbars, and sidebars.

You can use tabs:

```r
mainPanel(
    tabsetPanel(type = 'tabs',
        tabPanel('tab1', br(), textOutput('out1'))
        tabPanel('tab2', br(), textOutput('out2'))
        tabPanel('tab3', br(), textOutput('out3'))
    )
)
```

# Interactive Graphics

You can create graphics that the user an interact with. One method is to select multiple data points on a graph by using the `brush` argument in `plotOutput()` on the server side.

On the UI side:

```r
mainPanel(
    plotOutput('plot1', brush = brushOpts(
        id = 'brush1'
    ))
)
```

On the server side:

```r
shinyServer(function(i, o) {
    model <- reactive({
        
        brushed_data <- brushedPoints(
            trees,
            i$brush1
            xvar = 'Girth',
            yvar = 'Volume'
        )
        
        if (nrow(brushed_data) < 2) {
            return(NULL)
        }
        
        return(lm(Volume ~ Girth, data = brushed_data))
    })
    
    ...
})
```

# Shiny Gadget

Shiny gadget is a function that launches a small, single page Shiny application. 

```{r}
library(shiny)
library(miniUI)

a_gadget <- function() {
    ui <- miniPage(
        gadgetTitleBar("A Gadget")
    )
    
    server <- function(i, o, session) {
        observeEvent(i$done, {
            stopApp()
        })
    }
    
    rungadget(ui, server)
}
```

One of the advantages of gadgets is that they are functions, so they can take arguments and can return values.

You can also do interactive graphics.

```r
pickTrees <- function() {
    ui <- miniPage(
        gadgetTitleBar('Select Points by Dragging your Mouse'),
        miniContentPanel(
            plotOutput('plot', height = '100%', brush = 'brush')
        )
    )

    server <- function(i, o, session) {
        o$plot <- renderPlot({
            plot(trees$Girth, trees$Volume, main = 'Trees!')
        })
        
        observeEvent(i$done, {
            stopApp(brushedPoints(trees, i$brush, xvar = 'Girth', yvar = 'Volume'))
        })
    }
        
    runGadget(ui, server)
}
```

# GoogleVis

Can create interactive charts:

```{r, results = 'asis'}
library(googleVis)

gvisMotionChart(Fruits, 'Fruit', 'Year')
```

There are a number of other charts:

- Motion charts: `gvisMotionChart()`
- Interactive maps: `gvisGeoChart()`
- Interactive tables: `gvisTable()`
- Line charts: `gvisLineChart()`
- Bar charts: `gvisColumnChart()`
- Tree maps: `gvisTreeMap()`

Combine multiple plots together:

```r
plot_a <- gvisGeoChart(...)
plot_b <- gvisTable(...)

merged <- gvisMerge(plot_a, plot_b, horizontal = FALSE)
```

# Plotly

Plotly is a web applications for creating and sharing data visualisations.

```{r}
library(plotly)
plot_ly(mtcars, x = ~wt, y = ~mpg, type = 'scatter', mode = 'markers')
```


Can change the colour of the points, size of the points

```{r}
mtcars %>% 
    plot_ly(x = ~wt, y = ~mpg, color = ~as.factor(cyl), size = ~hp)
```

Three dimensional scatterplot:

```{r}
tibble(
    x = 1:100,
    y = rnorm(100, 20, 10),
    z = rnorm(100, 20, 2),
    c = as.factor(sample(1:4, 100, replace = TRUE))
) %>% 
    plot_ly(x = ~x, y = ~y, z = ~z, type = 'scatter3d', mode = 'markers', color = ~c)
    
```

## Histogram

```{r}
tibble(
    x = rnorm(3000)
) %>% 
    plot_ly(x = ~x, type = 'histogram')
```

## Boxplot

```{r}
iris %>% 
    plot_ly(y = ~Petal.Length, color = ~Species, type = 'box')
```

## Heatmap

```{r}
terrain_d <- matrix(rnorm(100 * 100), nrow = 100)

plot_ly(z = terrain_d, type = 'heatmap')
```

## 3D Surface

```{r}
# Using the terrain from before
plot_ly(z = terrain_d, type = 'surface')
```

## Choropleth

```{r}
library(stringr)
library(glue)

df <- read_csv(
    "https://raw.githubusercontent.com/plotly/datasets/master/2011_us_ag_exports.csv"
)

map_options <- list(
    scope = 'usa',
    projection = list(type = 'albers usa'),
    showlakes = TRUE,
    lakecolor = toRGB('white')
)

df %>% 
    mutate(hover = glue("State: {code}")) %>% 
    plot_ly(
        type = 'choropleth',
        locations = ~code,
        locationmode = 'USA-states',
        z = ~`total exports`,
        colors = 'Blues',
        text = ~hover
    ) %>% 
    layout(title = 'USA Map', geo = map_options)
```

## GGPlotly

```{r}
(
    mtcars %>% 
    ggplot(aes(wt, mpg)) +
    geom_point() +
    geom_smooth(method = 'lm') +
    facet_wrap(~cyl)
) %>% 
    ggplotly()
            
```

