---
title: "Course 9 - Developing Data Products - Week 2 - Notes"
author: "Greg Foletta"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
    toc: true
---

# R Markdown Presentations

A new slide is a level 2 heading "##".

To do a slide without a title, start with '***'.

When running R code, you can set `comment = ""` to set the comment character or remove it completely:

```{r}
head(mtcars)
```

```{r, comment = ""}
head(mtcars)
```


# Leaflet

Allows for the creation of interactive maps. 

```{r}
library(leaflet)
leaflet() %>% 
    addTiles()
```

The mapping data is from OpenStreeMap.

We can add a marker:

```{r}
leaflet() %>% 
    addTiles() %>% 
    addMarkers(lat = -37.7298, lng = 144.9553, popup = 'Coburg Velodrome')
```

You can add HTML into the popups, e.g. anchors/links.

To add many markers, you can create a data frame:

```{r}
tibble(
    lat = -c(36, 37, 38),
    lng = c(143, 144, 145)
) %>% 
    leaflet() %>% 
    addTiles() %>% 
    addMarkers()

```

## Custom Markers

```{r}
ccc_icon <- makeIcon(
    iconUrl = 'https://scontent.fmel7-1.fna.fbcdn.net/v/t31.0-1/cp0/c7.0.56.56a/p56x56/12045418_1173078239386251_5340026945367359072_o.jpg?_nc_cat=106&_nc_sid=dbb9e7&_nc_ohc=CVP7qbMvVsoAX_B5sru&_nc_ht=scontent.fmel7-1.fna&oh=e8ba7d82868ac3198b6a2771a32ecc39&oe=5EA97D59'
)

leaflet() %>% 
    addTiles() %>% 
    addMarkers(lat = -37.7298, lng = 144.9553, popup = 'Coburg Velodrome', icon = ccc_icon)
```

## Mapping Clusters

```{r}
tibble(
    lat = runif(500, min = -37.8, max = -37.7),
    lng = runif(500, min = 144.8, max = 145)
) %>% 
    leaflet() %>% 
    addTiles() %>% 
    addMarkers(clusterOptions = markerClusterOptions())
```

You can add circle markers:

```{r}
tibble(
    lat = runif(500, min = -37.8, max = -37.7),
    lng = runif(500, min = 144.8, max = 145)
) %>% 
    leaflet() %>% 
    addTiles() %>% 
    addCircleMarkers()
```

Can use a variable in the data frame to determine the radius.

## Legend

Can add a legend with `addLegend()`.


