---
title: "Course 9 - Developing Data Products- Week 3 - Notes"
author: "Greg Foletta"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
    toc: true
---


```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
```

# R Package

- Mechanism for extending the basic functionality of R.
- Collection of R function or other data objects.
- Organised in a systematic funciton.


## Where?

- CRAN and Bioconductor.
- GitHub, BitBucket, etc

Packages from CRAN are installed with `install.packages()`, from GitHub with `install_guthub()` from `devtools`.

## Advantages

- Documentation / vignettes
- Centralised resources
- Minimal standards for reliability and robustness
- Maintainability / extension
- Interface definition / API
- Users know that it will at least load properly

## Process

- Write some code in an R script file.
- Incporate it into the package structure.
- Write documentation for user functions.
- Include other material (examples, demos, datasets, tutorials).
- Package it up.
- Submit to CRAN or Bioconductor.
- Push source code to GitHub or other repo site.
- People find problems
    - They tell you about it.
    - They fix the problem themselves.

## Contents

It's started by creating a directory with the name of the R package.

### DESCRIPTION

The DESCRIPTION file has:

- Package: aname of the package.
- Title: full, longer name
- Description: longer description of the package.
- Verion: version number, usually in M.m-p format
- Author, Authors@R: name of the original author(s).
- Maintainer: name and email of the person who fixes problems.
- License: license for the source code.

Other optional fields:

- Depends: R packages that your package depends on.
- Suggests: optional R packages that users may want to have installed.
- Date: Release date in YYYY-MM-DD format.
- URL: package home page
- Other: fields can be added.

### R Code

Copy the code into the **R/** subdirectory. There can be any number of files in this directory. Code for all functions should be here, nowhere else.

### NAMESPACE

This is used to indicate which functions are exported. Exported functions can be called by the user and are considered the public API. Non-exported functions cannot be called directly, but the code can be inspected. 

This hides the implementation details from users and makes for a clean package interface.

You can also indicate what functions you **import** from other packages. This allows for your package to use other packages without making other packages visible to the user.

Importing a function loads the package but does not attach it to the search list.

**Key directives:**

`export(<function>)`
`import(function)`
`importFrom(<package>, <function>)`

Also important:

`exportClasses(<class>)`
`exportMethods(<generic>)`

Example from `magrittr`:

```r
S3method("[",fseq)
S3method("[[",fseq)
S3method(print,fseq)
export("%$%")
export("%<>%")
export("%>%")
export("%T>%")
export("n'est pas")
export(add)
export(and)
export(debug_fseq)
export(debug_pipe)
export(divide_by)
```

### Documentation

Documentation files are *.Rd* files placed in the **man/** subdirectory. They're written in a specific markup language, and are required for every exported function.

You can document other things such as concepts and package overview.

Markup example:

```
name{\%>\%}
\alias{\%>\%}
\title{Pipe}
\usage{
lhs \%>\% rhs
}
\arguments{
\item{lhs}{A value or the magrittr placeholder.}
```

## Building and Checking

`R CMD build` is a command-line program that creates a package archive file (.tar.gz).

`R CMD check` runs a battery of tests on the package:

- Documentation exists.
- Code can be loaded, no major coding problems or errors.
- Run examples in documentation.
- Check docs match code.
- All tests must pass to put package on CRAN.

## Getting Started

The `package.skeleton()` function creates a skeleton R package. Puts the directory structure (R/, man/), DESCRIPTION, NAMESPACE, and documentation files.

If there are functions visible in your workspace, it writes the code files to the R/ directory, and documentation stubs are created in man/.

# R Classes and Methods

R is different in that it is both interactive **and** has a system for object orientation. Object orientation is a bit different in R that most other languages.

There are two styles of classes and methods:

- **S3**
    - Included with version 3 of the S language.
    - Informal, slightly 'kludgey'.
    - Sometimes called old-style.
- **S4**
    - More formal and rigorous.
    - Included with S-PLUS 6 and R 1.4.0 (December 2001).
    - Also called new-style.

The code for implementing S4 classes/methods is in the `methods` package.

## OO in R

A class can be defined using `setClass()` in the `methods` package. An object is an instance of a class. These are created using `new()`.

A generic function is an R function which dispatches methods. It usually encapsulates a 'generic' concept such as `predict()`. The generic doesn't actually compute anything.

## Classes

All objects in R have a class:

```{r, comment = ''}
class(1)

class(rnorm(100))

class(TRUE)

class('foo')
```

## S3 Generics

As an example, `mean()` and `print()` are generic:

```{r}
mean
print
```

You can see what methods are associated with a generic:

```{r}
methods('mean')
```

## S4 Generics

The `show()` function from `methods` is an S4 generic:

```{r}
show
```

See the methods associated:

```{r}
showMethods('show')
```

## Method

THe first argument for a generic function is an object of a particular class. The function checks the class of the object and searches to see if there is an appropriate dispatch method for the class.

If there is, the method is called on the object.

If there is no dispatch method, a search is done to see if there is a default method for the generic. If a default exists, it is called.

If there isn't a default, an error is thrown.

## Examining Code

You cannot jus print the code for a method. If you want to see the code for an S3 method you use `getS3method(<generic>, <class>)`. For S4 you use `getMethod(<generic>, <signature>)`.


```{r}
head(getS3method('mean', 'default'), 10)
```

Some methods are visible (`mean.default()`), but you shouldn't call these directly.

## Writing

Typically you'll write new methods for `print()`, `summary()` and `plot()`.

Two ways:

- Write a methods for the new class but for an existing generic function.
- Write new generic functions and new methods for those generics.

## S4 Classes

Defined using the `setClass()` function. You then define methods for the class with the `setMethod()` function.

```{r}
library(methods)
setClass(
    'polygon',
    representation(x = 'numeric', y = 'numeric')
)
```

A plot method can be created with the `setMethod()` function. You specify a generic and a signature. The sginature is a character vector which indicates the classes of objects accepted by the method.

```{r}
setMethod(
    'plot', 'polygon',
    function(x, y, ...) {
        plot(x@x, x@y, type = 'n', ...)
        xp <- c(x@x, x@x[1])
        yp <- c(x@y, x@y[1])
        lines(xp, yp)
    }
)
```

```{r}
a <- new('polygon', x = c(1,2,3,4), y = c(1,2,3,1))
plot(a)
```
