---
title       : 'Developing Data Products'
subtitle    : 'Course Assignment'
author      : 'Greg Foletta'
job         : 'Managing Consultant, Telstra'
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## Introduction

Shiny applications are a great way of allowing people to interact with data. Users can use sliders, boxes and other
types of input to manipulate and generate data.

This opens up a world of possibilities for the educational space, creating dynamic pages that allow students to enhance 
their learning opportunities.

---

## What Are We Solving

A student's first introduction to the world of regressions is a linear regression. This can often be an overwhelming experience
with a whole host of new terms and concepts introduced:

- Means
- Standard Deviations
- Confidence Intervals

How can we, interactively, introduce these concepts to a student?

---

## The Solution

**Shiny Applications!**

We have developed a Shiny application that allows a student to interact with a linear regression.

Using the Shiny application, the student can change parameters and get instance feedback on how this
changes the linear regression.

The student can change the following parameters:

- Mean of the $\epsilon$ error term.
- Standard deviation of the $\epsilon$ error term.
- The sample size.

---

## Example A

The following is an example of the output the student is initially shown, with sample size of 100, a mean of 10 and a standard deviation of 100:

```{r echo = FALSE, warning = FALSE}
tibble(
    x = 1:100,
    y = x + rnorm(100, mean = 10, sd = 100)
) %>% 
    ggplot(aes(x, y)) +
    geom_point() +
    geom_smooth(method = 'lm') +
    geom_abline(intercept = 0, slope = 1, colour = 'red') +
    labs(
        x = 'X Axis',
        y = 'Y',
        colour = '# Cylinders'
    )
```


---

## Example B

The following is an example of the output the student can manually change, with sample size of 1000, a mean of 50 and a standard deviation of 400:

```{r echo = FALSE, warning = FALSE}
tibble(
    x = 1:1000,
    y = x + rnorm(1000, mean = 50, sd = 400)
) %>% 
    ggplot(aes(x, y)) +
    geom_point() +
    geom_smooth(method = 'lm') +
    geom_abline(intercept = 0, slope = 1, colour = 'red') +
    labs(
        x = 'X Axis',
        y = 'Y',
        colour = '# Cylinders'
    )
```


---

## Thank You

