---
title: "Developing Data Products"
subtitle: "Week 3 Assignment"
author: "Greg Foletta"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(glue)
library(openxlsx)
library(plotly)
library(lubridate)
library(knitr)
library(leaflet)
```

```{r cache = TRUE}
FINAL_CLEANUP <- TRUE

bike_data_file <- 'cycle_data.zip'

download.file(
        url = 'https://vicroadsopendatastorehouse.vicroads.vic.gov.au/opendata/Traffic_Measurement/Bicycle_Volume_and_Speed/Bicycle_Volume_Speed_2019.zip',
        destfile = bike_data_file
    )


# Unzip the outer file
zip_files <- unzip(zipfile = bike_data_file, list = TRUE)
unzip(bike_data_file)

# Unzip the inner files
ex_dir <- 'csv_data'

if (!dir.exists(ex_dir)) {
    dir.create(ex_dir)
}

walk(zip_files$Name, ~unzip(.x, exdir = ex_dir))
```


```{r, message = FALSE, cache = TRUE}
# Read all of the files in
list.files(ex_dir, full.names = TRUE) %>% 
    map(~read_csv(.x)) %>% 
    bind_rows() %>% 
    mutate(
        DATETIME = as.POSIXct(paste(DATE, TIME), format = '%d/%m/%Y %H:%M:%S'),
        DATE = as.Date(DATE),
        SITE_XN_ROUTE = as.factor(SITE_XN_ROUTE)
    ) ->
    bicycle_d
```

```{r}
# Counter site listing
read.xlsx(
    'https://vicroadsopendatastorehouse.vicroads.vic.gov.au/opendata/Traffic_Measurement/Bicycle_Volume_and_Speed/VicRoads_Bike_Site_Number_Listing.xlsx',
    sheet = 'VR BIKE Counter Site Locations'
) %>% 
    select(LAT, LONG, LOC_LEG, LOC.DESC) %>% 
    mutate(
        type = str_extract(LOC.DESC, '^\\(.*?\\)')
    ) ->
    sites

# Add in the location description
bicycle_d %>% 
    left_join(sites, by = 'LOC_LEG') %>% 
    mutate(LOC_LEG = as.factor(LOC_LEG)) ->
bicycle_d
```


## Introduction

- In this presentation we look at bike usage at different points within Melbourne, Victoria.
- Data is from [discover.data.vic.gov.au](https://discover.data.vic.gov.au/dataset/bicycle-volume-and-speed).

## Site Locations - Map


```{r warning = FALSE}
sites %>% 
    leaflet() %>% 
    addTiles() %>% 
    addMarkers(lat = ~LAT, lng = ~LONG, popup = ~LOC.DESC)
```



## Top Sites

- Bike usage is determined by a counter at different locations.
- We see a large differential between one site.

## Top Sites - Figure

```{r}
bicycle_d %>% 
    count(LOC_LEG) %>% 
    top_n(10) %>% 
    pull(LOC_LEG) -> 
top_10_legs

(
bicycle_d %>% 
    filter(LOC_LEG %in% top_10_legs) %>% 
    count(LOC_LEG) %>% 
    ggplot() +
    geom_col(aes(fct_reorder(LOC_LEG, n), n)) +
    coord_flip() +
    labs(
        x = 'Top 10 Locations',
        y = 'Total Count of Bikes'
    )
) %>% 
    ggplotly()
```


## Bikes Per Day

- The following graph shows bikes per day over the year for the the top 10 sites.
- It appears there is incomplete data for everything but the top reporting site.

## Bikes Per Day - Digure

```{r}
(
bicycle_d %>%
    filter(LOC_LEG %in% top_10_legs) %>% 
    group_by(LOC_LEG) %>% 
    count(DATE) %>% 
    mutate(DATE = as.POSIXct(DATE)) %>% 
    ggplot() +
    geom_line(aes(DATE, n, colour = fct_reorder(LOC_LEG, desc(n)))) +
    labs(
        x = 'Date',
        y = 'Number of Bikes',
        colour = 'Location'
    )
) %>% 
    ggplotly()
```

## Top Location

- Where is the 60182 location?
```{r comment = "" }
sites %>% 
    dplyr::filter(LOC_LEG == 60182) %>% 
    slice(1) ->
    black_rock


leaflet() %>% 
    addTiles() %>% 
    addMarkers(lat = black_rock$LAT, lng = black_rock$LONG, popup = black_rock$LOC.DESC)
```

## Black Rock - Intra-day Counts

- The following graph shows intra-day count per hour, per month.

```{r}
(
bicycle_d %>% 
    dplyr::filter(LOC_LEG == 60182) %>% 
    mutate(
        hour = hour(TIME),
        month = month(DATE, label = TRUE)
    ) %>% 
    count(month, hour) %>% 
    ggplot() +
    geom_line(aes(hour, n, colour = month)) +
    labs(
        x = 'Hour of the Day',
        y = 'Total Bike Count',
        colour = 'Month'
    )
) %>% 
    ggplotly()

    
```

```{r}
# Cleanup
if (FINAL_CLEANUP) {
    file.remove(Sys.glob('*.zip'))
    unlink('csv_data', recursive = TRUE)
    unlink('index_cache', recursive = TRUE)
}
```

