---
title: "Course 2 - R Programming - Week 1 - Notes"
author: "Greg Foletta"
output:
  html_document:
    toc: true
    theme: cerulean
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

# Pre-Lecture - Writing Code

Working directory found with `getwd()`.

See the files in a directory with `dir()`.

# Overview and History of R

R is a dialect of the S language.

S was developed by John Chambers and others at Bell Labs.

R was developed in 1991 in New Zealand by Ross Ihaka and Robert Gentleman.

R version 1 released in 2000.

# Features of R

- Syntax is similar to S-PLUS
- Semantics are superficially like S, but are in reality quite different.
- Runs on almost any standard playform/OS
- Frequent releases.
- Quite lean, functionality in modular packages.
- Graphics capabilities are very sophisticated and better than most stat packages.
- Useful for interactive work.
- Active and vibrant user community.

# Free Software

With **free** software you are granted:

- The freedom to run the program.
- The freedom to study how the program works and apapt it to your needs.
- The freedom to redistribute copies.
- The freedom to improve the program.

# Drawbacks of R

- Essentially based on 40 year old technology.
- Little built in support for dynamic or 3D graphics.
- Functionality is based on consumer demand and user contributions.
- Objects must generally be stored in physical memory - however advancements have been made on this.
- Not ideal for all possible situations.

# Design

1. Base R
1. Everything else

Base R comes with base packages (stat, util, etc) as well as recommended packages.

# Data Types - R Objects and Attributes

Everything is R is an object.

R has five basic "atomic" classes of objects:

- Character
- Numeric (real numbers)
- Integer
- Complex
- Logical

The most basic object is a vector. Can only contain objects of the same class. Empty vectors are created with the `vector()` function.

## Numbers

Generally treated as numberic objects (double precision real numbers. Need to provide the `L` suffix if you want an integer.

```{r numbers}
real_vec <- c(1,2,3,4)
int_vec <- c(1L, 2L, 3L, 4L)

real_vec %>% class()
int_vec %>% class()
```

`Inf` represents infinity. `NaN` represents an undefined value.

## Attributes

R objects can have attributes. `names` and `dimnames` are common - for example a matrix has the number of rows and columns:

```{r attributes}
example_attr <- matrix(1:16, ncol = 4)
example_attr

print('Attributes:')
example_attr %>% attributes()
```

You also have `class`, `length`, and other user defined attributes / metadata.

# Vectors and Lists

`c()` function can be used to create vectors, as well as `vector()`.

```{r vectors}
c(1:10)
vector('numeric', 10)
```

If you mix the classes, values are coerced you'll get the lowest common denominator.

```{r mixing}
c(1,2,3L, "a") %>% class()
c(TRUE, 1, 2) %>% class()
c(TRUE, 1, "a") %>% class()
```

## Explicit Coercion

Can use explicit coercion:

```{r explicit_coercion}
c(1,2,3) %>% as.character() %>% class()
```
Nonsensical coercion will result in `NAs`.

```{r nonsense_coercion}
c("1", "2") %>% as.logical()
```

## Lists

Special type of vector that can contain elements of different classes.

```{r list}
example_list <- list("a", c(1:10), TRUE)
example_list
```

Using a single bracket on a list returns a list of one. Using a double bracket returns the element in the list

```{r list_index}
example_list[1]
example_list[1] %>% class()

example_list[[1]]
example_list[[1]] %>% class()
```

# Matrices

A special type of vector with a 'dimension' attribute. The atrribute itself is a vector of length two (rows / cols).

```{r matrix}
matrix(1:16, nrow = 4)
```

Matrices are built column wise - see above.

Can be created from vectors:

```{r vec_to_matrix}
mv <- c(1:16)
mv

# Assign the attribute
attr(mv, 'dim') <- c(4,4)
mv

# Use the dim() function
dim(mv) <- c(2,8)
mv
```

# Column and Row Binding

Can column or row bind a vector:

```{r binding}
cbind(c(0:7), c(32:39))
rbind(1:4, 8:11)
```
# Factors

Special vector used to represent categorical data. There's ordered and unordered factors.

Can be thought of as an integer vector with labels.

```{r factor}
example_factor <- factor(c('y', 'y', 'n', 'n', 'y'))
example_factor

example_factor %>% class()
example_factor %>% typeof()

unclass(example_factor)
```
## Factor Ordering

Can be set using the `levels` argument to `factor()`. The levels are set alphabetically - hence why 'n' above was coded as 1.

```{r factors_ordered}
factor(c('y', 'y', 'n', 'n', 'y'), levels = c('y', 'n')) %>% unclass()
```

# Missing Values

Denoted by `NA` or `NaN`. Use `is.na()` or `is.nan()` to determine if an object of of that value.

The NAs can have a class as well.

An NaN is also considered NA.

```{r na}
na_vec <- c(1,2,NaN, NA, 4)
is.na(na_vec)
is.nan(na_vec)
extract <- na_vec[4]
extract
extract %>% class()
```

# Data Frames

Special type of list where every element has the same length. Each element can be considered a column. Different classes in each column.

They have an attribute `row.names`.

# Names Attribute

Objects can have names.

```{r named_vector}
named_vec <- c(1,2,3)
named_vec
names(named_vec) <- c('One', 'Two', 'Three')
named_vec
```

Named lists:

```{r named_list}
named_list <- list(one = 1, two = 2, three = 3)
named_list
```

Matrices have dimnames:

```{r named_matrices}
named_matrix <- matrix(1:4, ncol = 2)

dimnames(named_matrix) <- list(c('A', 'B'), c('C', 'D'))

named_matrix
```

# Reading Tabular Data

- `read.table()` and `read.csv()` for reading tabular data.
` `readLines()` for reading lines of a text file.
- `source()` for reading R code.
- `dget()` for reading deparsed R objects.
- `load()` for workspaces.
- `unserialize()` for reading single R objects in binary form.

## Read Table

Arguments:

- `colClasses` - character vector indicating the class of each column.
- `comment.char` - character string indicating the comment character.
- `stringsAsFactors` - should character variables be coded as factors?
    - Defaults to `TRUE`.

Read table generally figures out the classes, skips lines beginning `#`, and figures out how many rows.

`read.csv()` is identical except the default separator is a comma instead of a space and expects a header line.

# Reading Large Tables

- Make a rough calculation on the memory required and determine if you have enough physical memory.
- Set `comment.char = ''` if there are no comments.

Using the `colClasses` argument can make reading in much faster, and setting `nrows` can help memory usage, with a mild overestimate being okay.

Know your system - memory, applicatons, other users, OS.

You can calculate memory requirements. Consider 1,000,000 rows and 200 columns of numeric data. Numeric data is 8 bytes.

```{r memcalc}
# Calculate Gigabytes of storage.
(1000000 * 200 * 8) / 2^30
```

# Textual Formats

`dump()` and `dput()` are useful. 

`dump()` dumps named objects from the environment into text format:
```{r dump}
dump_obj <- tibble(x = 1:4, y = 2:5)
vec <- c(1:10)

dump(c('dump_obj', 'vec'), file = tempfile())
```
File has the following contents:
```r
dump_obj <-
structure(list(x = 1:4, y = 2:5), row.names = c(NA, -4L), class = c("tbl_df", 
"tbl", "data.frame"))
vec <-
1:10
```

`Can use `dput()` dumps a single object:

```{r dput}
tibble(x = 1:16, blah = letters[x]) %>%
    dput(file = tempfile())
```

File contains:
```r
structure(list(x = 1:16, blah = c("a", "b", "c", "d", "e", "f", 
"g", "h", "i", "j", "k", "l", "m", "n", "o", "p")), row.names = c(NA, 
-16L), class = c("tbl_df", "tbl", "data.frame"))
```

Textual formats are good with version control, easier to fix 'corruption', and adhere to the Unix philosophy. However they are space-inefficient.

# Connections

Connections can be made to:

- `file()`
- `gzfile()` - gzip compressed file.
- `bzfile()` - bzip2 compressed file.
- `url()`

# Subsetting

The `[` returns the same class as the original, can be used to extract more than one object (one exception).

```{r subset}
x <- c(1:10)
x[4]
x > 5
x[x > 5]
```

The `[[` returns the objects within.

The $ can extract by name, and the semantics are similar to `[[`.

## Lists

```{r subset_list}
l <- list(a = 1:4, b = letters[1:4], c = 80:83)

l[c(1,3)]

name <- 'a'
l[name]
l[[name]]
```

## Nested Elements

```{r nested_list}
l <- list(
    a = list(
        b = "Nest B",
        c = "Nest C"
    ),
    d = "Element D",
    e = "Element E"
)

# Access the nested element
l[[c(1,2)]]
l[[1]][[1]]
```

## Matrices

Can be subsetted in the normal way using $i,j$ row / column indicies.

```{r matrix_subset}
m <- matrix(1:16, ncol = 4)

# 2nd Row
m[2,]
# First column
m[,1]
# Single element
m[1,4]
```

Subsetting out a matrix is a vector, not another matrix.

If you want a matrix you need to use the `drop` argument:

```{r matrix_drop}
# 2nd Row
m[2,,drop = F]
# First column
m[,1, drop = F]
# Single element
m[1,4, drop = F]
```

## Partial Matching

Works with the `$` and the `[[`. Double bracket expects exact, but can use `exact = F` argument,

## Removing Missing Values
Use `is.na()` or `complete.cases()`.

```{r remove_na}
x <- c(1,2,3,NA,4,NA,NA,5)
x[!is.na(x)]
x[complete.cases(x)]
```

Can be done on data frames too - done by rows.

```{r remove_na_df}
x <- data.frame(
    a = c(1,2,NA,NA,5,6),
    b = c(1,NA,3,4,5,6)
)

complete.cases(x)

x[complete.cases(x),]
```
