---
title: "Course 2 - R Programming - Week 2 - Notes"
author: "Greg Foletta"
output:
  html_document:
    toc: true
    theme: cerulean
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Control Structures

Allow you to control the flow of a program.

# If-Else

Standard if-else:

```{r if_else}
if (rbinom(1,1,.5)) {
    print("True")
} else {
    print("False")
}
```

You can also assign from the construct:

```{r if_else_assign}
y <- if(rbinom(1,1,.5)) {
    10
} else {
    20
}

y
```

# For

Standard for:

```{r for}
for (i in 1:3) {
    print(i)
}
```

Can use `seq_along()` to get integer vector:

```{r for_seq_along}
vec <- LETTERS[1:6]

for (i in seq_along(vec)) {
    print( paste(i, LETTERS[i]) )
}
```

# While Loops

Standard while loop:

```{r while}
i = 4
while (i) {
    print(i)
    i = i - 1
}
```

# Repeat, Next, Break

Repeat is an infinite loop:

```{r repeat}
i = 0
repeat {
    i = i + 1
    print(i)
    
    if (i == 1) {
        next
    }
    
    if (i == 4) {
        break
    }
}
```

# Functions

Functions are first class - treated like everything else.

Functons can be nested, so you can define a function inside another function.

## Function Arguments

**Formal** arguments are the arguements included in the function definition. `formals()` function returns a list of format arguments.

```{r formal}
formals(mean)
```
Function arguments can be *missing* or might have default values.

Matched positionally or by name. When an argument is matched by name, it is 'taken out', and the remaining arguments are matched in their order.

Function arguments **can** be partially matched. Checks for exact match, then partial match, then positional match.

## Lazy Evaluation

Arguments are only evaluated as needed.

```{r func_def}
func <- function(a, b) {
    a^2
}

func(2)
```

No error not specifying `b` as it's not used.

## The '...' Argument.

Indicates a variable number of arguments often passed to other functions.

```{r dots}
my_mean <- function(x, ...) {
    mean(x, ...)
}

v <- c(1,2,3,NA)
my_mean(v)
my_mean(v, na.rm = T) 
```

Also used when the number of arguments can't be known:

```{r args_dots}
args(paste)
```

Of course args after the dots must be named and cannot be partially matched.

# Scoping Rules

R searches through environments to find a symbol:

```{r search}
search()
```

The global environment is searched first, with all other packages searched in the order of the `search()` list.

When a user loads a package with `library()`, the namespace (environment) of the package is placed into position 2.

R has separate namespaces for functions and non-functions, so it's possible to have an object and a function named `c`.

R uses *lexical* scoping or *static scopring*.

Consider the following function:

```{r lexical}
f <- function(a, b) {
    a + b * z
}
```

`z` is a *free variable*. What value is assigned to `z`?

**Lexical Scoping** - variables are seaerched for in the environment in which the function is defined.

If it's not found, the search is continued in the *parent environment*.

After the top level environment, the search continues down the search list until we hit the empty environment.

## Lexical vs Dynamic

With **lexical** scoping, the value is looked up in the environment in which the function was defined.

With **dynamic** scoping, the value is looked up in the environment from which the function was *called*, the calling environment.

Often functions are defined and called in the global environment, making it seem as if the scoping is dynamic.

## Lexical Consequences

All kfunctions must carry a pointer to their respective defining environments, which could be anywere.

## Scoping Optimisation

There are some statistical optimisation functions like `optim()`, `nlm()` and `optimize()` that you pass a function to.

However an object may depend on a host of other things besides its parameters - like data.

We write a constructor function which constructs the objective function.

Close over the data you want to carry around, then call the constructed function with the variable parameters.

```{r closures}
make.func <- function(data, fixed = c(F, F)) {
    params <- fixed
    function(p) {
        params[!fixed] <- p
        params
    }
}

gen1 <- make.func(c(1,2,3,4))

gen1
ls(environment(gen1))
```

# Coding Standards

1. Always use text files / text editor.
1. Indent your code.
    - Can use Ctrl + I in RStudio
1. Limit the width of your code (80 columns)
1. Limit the length of individual functions.

# Dates and Times

Dates are represented by the `Date` class, times by the `POSIXct` class.

Dates are stored as the number of days since 1970-01-01, times the number of seconds since 1970-01-01.

```{r date}
x <- as.Date('1984-04-27')
unclass(x)

weekdays(x)
months(x)
quarters(x)
```

`POSIXct` is just a very large integer under the hood.

`POSIXlt` is a list underneath that stores other useful information.

```{r time}
x <- Sys.time()
x
class(x)
weekdays(x)
unclass(x)
```

Use `strptime()` to create dates in a different format.


```{r dt_operations}
y <- as.POSIXct(as.Date('1990-01-01'))
y < x
x - y
```

Takes account of leap years, leap seconds, and timezones.

```{r timezone}
as.POSIXct('1990-01-01 11:00:00') - as.POSIXct('1990-01-01 11:00:00')

as.POSIXct('1990-01-01 11:00:00') - as.POSIXct('1990-01-01 11:00:00', tz = 'AEST')
```
