---
title: "Course 2 - R Programming - Week 3 - Notes"
author: "Greg Foletta"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Loop Functions

- `lapply()` - functionally loop over a list.
- `sapply()` - same as `lapply()` but try to simplify the result.
- `apply()` - apply a function over the margins of an array.
- `tapply()` - apply a function over subsets of a vector.
- `mapply()` - multivariate version of `lapply()`.

`split()` is also useful in conjunction with `lapply()`.

# lapply

Takes a list, a function, and ... which are passed to the function.

If the first argument is not a list, it is coerced into a list.
```{r lapply}
val <- list(a = 1:20, b = rnorm(100))
lapply(val, mean)
lapply(1:4, rnorm, mean = 100, sd = 3)
```

# sapply

Will try to return to simplify the result of `lapply()` if possible.

- If the result is a list where every length is 1, a vector is returned.
- If the result is a list where every element is a vector of the same length (> 1), a matrix is returned.
- Otherwise a list is returned.

```{r sapply}
ex_list <- list(a = 0:3, b = 8:11)
sapply(ex_list, mean)
sapply(ex_list, function(x) x + 1)
```

# apply

Used to evaluate a function over the margins of an array.

- Most often used to apply a function to the rows or columns of a matrix.
- Can be used with general arrays e.g. taking the average of an array of matrices.
- Not faster than a loop, but is one line.

```{r apply}
ex_mat <- matrix(rnorm(8), ncol = 2, nrow = 4)

# Mean of the columns
apply(ex_mat, 1, mean)

# Mean of the rows
apply(ex_mat, 2, mean)

# Mean of the rows and columns
apply(ex_mat, c(1,2), mean)

```

The `c(1,2)` would be used if it was multi-dimensional matrix.

```{r apply_3d}
ex_array <- array(rnorm(2 * 2 * 4), dim = c(2, 2, 4))
apply(ex_array, c(1,2), mean)
```

# mapply

Multivariate apply - applies a function in parallel over a set of arguments.

```{r mapply}
mapply(rep, 1:4, c(2,6))
```

Can be used to vectorise functions.

# tapply

Apply functions over subset of a vector.

```{r tapply}
x <- c(rnorm(2000), rnorm(2000, 10), rnorm(2000, 30))
levels <- gl(3, 2000)
tapply(x, levels, mean)
```
# split

Takes a vector or other objects and splits into groups determined by a factor or list of factors.

```{r split}
ex_vector <- 1:16
levels <- gl(4,4)
split(ex_vector, levels)
```

Can use `lapply()` or `sapply()` with the split.

```{r split_lapply}
lapply(split(ex_vector, levels), mean)
```

Can use it on data frames.

```{r split_df}
# Split by the month
aq_monthly <- split(airquality, airquality$Month)
lapply(
    aq_monthly,
    function(x) colMeans( x[, c('Ozone', 'Solar.R')], na.rm = T )
)
```

Can use `sapply()` - as the returned values are the same length, you can get the results in a matrix:

```{r split_sapply}
aq_monthly <- split(airquality, airquality$Month)
sapply(
    aq_monthly,
    function(x) colMeans( x[, c('Ozone', 'Solar.R')], na.rm = T )
)
```

You can split on more than one level - e.g. male / female and another eye colour. You can use `drop = T` to drop the empty levels.

```{r split_levels}
x <- 1:18
gender <- gl(2, 3)
eye_colour <- gl(3, 2)
interaction(gender, eye_colour)

split(x, list(gender, eye_colour))
```

# Debugging

- `message()`: generic notification
- `warning()`: indication someting is wrong but not fatal.
- `error()`: indication a fatal error has ocurred.
- `condition` generic concept for indicating something unexpected can occur.

**Aside**: To return from a function without printing, can use `invisible()`.

## Tools

- `traceback()` prints the function call stack after an error ocurrs.
- `debug()` - flags a function for "debug" mode which allows you to step through one line at a time.
- `browser()` suspends the execution and moves into debug mode.
- `trace()` allows you to insert debugging code into a function.
- `recover()` - allows you to modify the error behaviour so that you can browse the function call stack.


